---
title: "Reproducible Data Programming Assignment 1"
author: "yiudim"
output: "html_document"
---

# Necessary work to get the data in...
## load necessary libraries
```{r}
require(dplyr)
require(ggplot2)
require(Hmisc)
```
	
## read the data file
```{r}
fileurl <- "./activity.csv"
pa1raw <- read.csv(fileurl)
```

## raw data pre-process -> make "date" into Date format
```{r}
pa1raw <- pa1raw %>% 
mutate(fdate = as.Date(date, format("%Y-%m-%d"))) %>% 
mutate(intervalt = as.POSIXct(formatC(interval, width=4, flag="0"),format="%H%M")) %>%
select(-date)
```
		

# What is the mean total of steps taken per day?
## calcuate total steps per day
```{r}
pa1 <- pa1raw %>% 
group_by(fdate) %>%
summarise(total_steps = sum(steps))
```

## show total steps per day in a histogram
```{r echo=FALSE}
hist(pa1$total_steps, xlab="Total Steps per Day", col="green", breaks=30)
```

## calculate and report mean and median of the total number of steps taken per day
```{r}
meanvalue <- as.integer(mean(pa1$total_steps,na.rm=TRUE))
medianvalue <- median(pa1$total_steps,na.rm=TRUE)
```

Mean is `r meanvalue`

Median is `r medianvalue`


# What is the daily average activity pattern?
## calcuate average steps per 5-minute interval
```{r}
pa1 <- pa1raw %>% 
group_by(intervalt) %>%
summarise(average_steps = mean(steps, na.rm=TRUE))
```

## plot average steps per 5-min interval in time series plot
```{r echo=FALSE}
with(pa1, plot(intervalt, average_steps, type="l", xlab="5-minute Intervals", 
main="Average Steps Taken per 5-min Intervals"))
abline(v=pa1$intervalt[which(pa1$average==max(pa1$average_steps))], col="blue")
```
Average steps per 5-min interval `r pa1$interval[which(pa1$average==max(pa1$average_steps))]`


# Imputing missing values 
## Calculate and report the total number of missing values in the dataset
How many NAs are there in the dataset (activities.csv) `r sum(is.na(pa1raw))`

## Assign average value of interval if step is NA
```{r}
pa1c <- pa1raw
for(i in 1:nrow(pa1c)) {
if(is.na(pa1c$steps[i])) {
pa1c$steps[i] <- as.integer(pa1$average_steps[which(pa1$intervalt == pa1c$intervalt[i])])
}
}
```

## Re-look at total, mean, and median values
```{r}
pa2 <- pa1c %>% 
group_by(fdate) %>%
summarise(total_steps = sum(steps))
```

## show total steps per day in a histogram
```{r echo=FALSE}
hist(pa2$total_steps, xlab="Total Steps per Day", col="green", breaks=30)
```

## calculate and report mean and median of the total number of steps taken per day
```{r}
Adjusted mean with imputed NAs is `r as.integer(mean(pa2$total_steps,na.rm=TRUE))
Adjusted mediam with imputed NAs is `r median(pa2$total_steps,na.rm=TRUE)`
```

## values are different from when we calcuated the same stats excluding the NAs in the dataset


# Activity patterns weekday vs weekend
```{r}
for(i in 1:nrow(pa1c)) {
dofw <- weekdays(pa1c$fdate[i]) 
if(dofw %in% c("Saturday","Sunday")){
pa1c$dayofweeks[i] <- "weekend"
} else {
pa1c$dayofweeks[i] <- "weekday"
}
}
```

## setup graphing panel
```{r echo=FALSE}
	par(mfrow=c(2,1))
	par(cex = 0.6)
	par(mar = c(0, 0, 0, 0), oma = c(4, 4, 0.5, 0.5))
```

```{r}
pa3 <- pa1c %>% 
filter(dayofweeks == "weekday") %>%
group_by(intervalt) %>%
summarise(average_steps = mean(steps, na.rm=TRUE))
```

## plot average steps per 5-min interval in time series plot
```{r echo=FALSE}
with(pa3, plot(intervalt, average_steps, type="l", xlab="5-minute Intervals")) 
title(main="Average Steps Taken per 5-min Intervals - Weekdays", line = -2)
abline(v=pa3$intervalt[which(pa3$average==max(pa3$average_steps))], col="blue")
```

5-min internal with max average number of steps (with imputed NAs) `r pa3$interval[which(pa3$average==max(pa3$average_steps))]`

## Remain Unchanged (with and without imputing NAs with average values
```{r}
pa4 <- pa1c %>% 
filter(dayofweeks == "weekend") %>%
group_by(intervalt) %>%
summarise(average_steps = mean(steps, na.rm=TRUE))
```

## Are there differences in activity patterns between weekdays and weekends?
```{r echo=FALSE}
with(pa4, plot(intervalt, average_steps, type="l", xlab="5-minute Intervals")) 
title(main="Average Steps Taken per 5-min Intervals - Weekends", line=-2)
abline(v=pa4$intervalt[which(pa4$average==max(pa4$average_steps))], col="blue")
```

pa4$interval[which(pa4$average==max(pa4$average_steps))]
## 915
